The role of cognateness in non-native spoken word recognition

### 2021-12-03 Rmd/manuscript.Rmd
Add manuscript plots with diagnostics

bibliography:
  - r-references.bib
  - references.bib
floatsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
csl: "apa7.csl"
documentclass: apa7
classoption: man
output:
  word_document:
    reference_docx: "manuscript_template.docx"
---

@@

Listening to speech in a non-native language is more costly than doing so in native languages, regardless of the level of proficiency [REF]. The source of this increased cognitive effort is likely to originate at multiple levels. For example, some sounds in the speech signal do not correspond to any phoneme in the native language, making the process of matching sounds to phonemes complicated. This is the case of a Spanish native listening to words containing the guttural /ʁ/ in French. Language learners, however, are rarely completely naïve to the language they are listening to. All languages share, to some extent, similarities, frequently due to their typological closeness. These similarities which can be exploited by non-native listeners to process the speech signal. In the present research we investigate the extent to which adult listeners rely on phonological similarity between cross-language synonyms during word translation.

@@

Cognates play a major role in virtually all models of bilingual lexical processing because they provide evidence that bilinguals access the lexicon in a language non-selective way. During word production and comprehension, lexical representations of both languages are activated, and the form similarity between both impacts bilinguals' performance in naming or lexical decision tasks [e.g., @costa2000cognate; @thierry2007brain]. In a seminal study, @costa2000cognate asked Spanish-Catalan adult bilinguals to name pictures in Spanish. Unbeknownst to participants, half of the labels associated to pictures were cognates in Spanish and Catalan (e.g., *puerta*-*porta*) while the other half were non-cognates (e.g., *mesa*-*taula*, meaning *table* in Spanish and Catalan). Surprisingly, Spanish-Catalan bilinguals named cognate pictures faster than non-cognate pictures, while Spanish monolinguals did not show this effect. This suggests that bilinguals activate the picture's word representations in both languages, and the phonological overlap between both labels facilitated the naming process.

Additionally, cognates seem to be learnt more easily in L2 than non-cognates. @de2000hard presented Dutch natives pairs of Dutch words and pseudowords. Participants' performance learnt and retained better pairs with high orthographic similarity than pairs with low orthographic similarity [see also @lotto1998effects for Dutch natives learning Italian words].

Cognates are not only learnt faster than non-cognates, but translated faster too. Early accounts of bilingual lexical access suggest that low-proficiency learners first established links between newly acquired words in L2 and their meanings through their L1 translation equivalents [@potter1984lexical]. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on the mediation of L1 representations. The Revised Hierarchical Model [RHM; @kroll1994category] captured this assumption and predicted that translating words from L1 to L2 (forward translation) should take longer than translating from L2 to L1 (backward translation). The rationale behind this prediction is that backward translation relies more strongly on direct word-word links between L1 and L2 representations, while forward translation would rely more strongly on the mediation between the concept and the two word forms. One of the consequences of this prediction is that backward translation should be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during backward translation.

@@

Since @kroll1994category proposed their RHM model, subsequent studies brought attention to the role of the network of connections that words establish with each other at the form-level (phonological or orthographic) or at the conceptual level. The Neighbourhood Activation Model [NAM; @luce1998recognizing] suggests that lexical selection is mediated not only by the lexical frequency of the target word (and the number of activated candidates), but also by the frequency of the candidates.

>>>>

bibliography:
  - r-references.bib
  - references.bib
floatsintext: yes
figurelist: no
tablelist: no
footnotelist: no
linenumbers: yes
mask: no
draft: no
csl: "apa7.csl"
documentclass: apa7
classoption: man
appendix: "manuscript_appendix.Rmd"
output:
  word_document:
    reference_docx: "manuscript_template.docx"
---

@@

Listening to speech in a non-native language is more costly than doing so in native languages, regardless of the level of proficiency [@lecumberri2010non]. The source of this increased cognitive effort is likely to originate at multiple levels. For example, some sounds in the speech signal do not correspond to any phoneme in the native language, making the process of matching sounds to phonemes complicated. This is the case of a Spanish native listening to words containing the guttural /ʁ/ in French. Language learners, however, are rarely completely naïve to the language they are listening to. All languages share, to some extent, similarities, frequently due to their typological closeness. These similarities which can be exploited by non-native listeners to process the speech signal. In the present research we investigate the extent to which adult listeners rely on phonological similarity between cross-language synonyms during word translation.

@@

Cognates play a major role in virtually all models of bilingual lexical processing because they provide evidence that bilinguals access the lexicon in a language non-selective way. During word production and comprehension, lexical representations of both languages are activated, and the form similarity between both impacts bilinguals' performance in naming or lexical decision tasks [e.g., @costa2000cognate; @thierry2007brain]. In a seminal study, @costa2000cognate asked Spanish-Catalan adult bilinguals to name pictures in Spanish. Unbeknownst to participants, half of the labels associated to pictures were cognates in Spanish and Catalan (e.g., *puerta*-*porta*) while the other half were non-cognates (e.g., *mesa*-*taula*, meaning *table* in Spanish and Catalan). Surprisingly, Spanish-Catalan bilinguals named cognate pictures faster than non-cognate pictures, while Spanish monolinguals did not show this effect. This suggests that bilinguals activate the picture's word representations in both languages, and the phonological overlap between both labels facilitated the naming process. Additionally, cognates seem to be learnt more easily in L2 than non-cognates. @de2000hard presented Dutch natives pairs of Dutch words and pseudowords. Participants' performance learnt and retained better pairs with high orthographic similarity than pairs with low orthographic similarity [see also @lotto1998effects for Dutch natives learning Italian words].

Cognates are not only learnt faster than non-cognates, but translated faster too. Early accounts of bilingual lexical access suggest that low-proficiency learners first established links between newly acquired words in L2 and their meanings through their L1 translation equivalents [@potter1984lexical]. As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on the mediation of L1 representations. The Revised Hierarchical Model [RHM,   @kroll1994category] captured this assumption and predicted that translating words from L1 to L2 (forward translation) should take longer than translating from L2 to L1 (backward translation). The rationale behind this prediction is that backward translation relies more strongly on direct word-word links between L1 and L2 representations, while forward translation would rely more strongly on the mediation between the concept and the two word forms. One of the consequences of this prediction is that backward translation should be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during backward translation.

@@

Since @kroll1994category proposed their RHM model, subsequent studies brought attention to the role of the network of connections that words establish with each other at the form-level (phonological or orthographic) or at the conceptual level. The Neighbourhood Activation Model [NAM, @luce1998recognizing] suggests that lexical selection is mediated not only by the lexical frequency of the target word (and the number of activated candidates), but also by the frequency of the candidates.

### 2021-12-03 Rmd/manuscript_appendix.Rmd
Add manuscript plots with diagnostics

```{r accuracy, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.cap="Proportion of correct translation by item. Words presented to participants in the English-Spanish or in the English-Catalan and Spanish-Catalan are listed in the Y-axis, ordered from higher to lower average translation accuracy, depicted in the X-axis. Dots and whiskers represent the average accuracy and 95% confidence interval of each word. Accuracy is plotted separately for each group."}

>>>>

```{r appendix-diagnostics, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

tar_load_globals()
tar_load(stimuli)
tar_load(participants)
tar_load(responses)
tar_load(model_fits)

bayesplot_theme_set(
  new = theme_custom() + 
    theme(
      axis.text.x = element_text(size = 8), 
      axis.text.y = element_text(size = 8),
      strip.text = element_text(size = 10),
      legend.key = element_rect(colour = NA, fill = "white")
    )
)

# Appendix 1: Stan code

```{r appendix-model-code, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
model_fits$fit_3$model
color_scheme_set(c("#1A85FF", "#9ccaff", "#1A85FF",  "#1A85FF", "#1A85FF", "#1A85FF"))

# Appendix 2: MCMC diagnostics

## Pairs plot

```{r appendix-mcmc-diagnostics, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
posterior_cp <- as.array(model_fits$fit_3)

plot_pairs <- mcmc_pairs(
  posterior_cp, pars = c("b_Intercept", "b_frequency_zipf", "b_pthn", "b_lv", "b_pthn:lv"),
  off_diag_args = list(size = 0.75),
  alpha = 1, size = 0
)

plot_pairs
```

## Trace plots

```{r appendix-trace-plot}
mcmc_trace(model_fits$fit_3, pars = c("b_Intercept", "b_frequency_zipf", "b_pthn", "b_lv", "b_pthn:lv"), n_warmup = 0) +
  annotate(geom = "rect", xmin = -Inf, xmax = 500, ymin = -Inf, ymax = Inf, alpha = 0.5, fill = "grey")

## Gelman-Rubin convergence criterion (R-hat)

```{r appendix-rhat, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
rhats <- rhat(model_fits$fit_3)
ratios <- neff_ratio(model_fits$fit_3)

mcmc_rhat(rhats) + 
  mcmc_neff(ratios) +
  plot_layout(guides = "keep", ncol = 1) &
  theme(
    legend.position = "right"
  )

# Appendix 3: by item accuracy

```{r appendix-accuracy, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.cap="Proportion of correct translation by item. Words presented to participants in the English-Spanish or in the English-Catalan and Spanish-Catalan are listed in the Y-axis, ordered from higher to lower average translation accuracy, depicted in the X-axis. Dots and whiskers represent the average accuracy and 95% confidence interval of each word. Accuracy is plotted separately for each group."}
